---
title: "Reproducible Research - Project 1"
author: "Mala Viswanath"
date: "Friday, July 17, 2015"
output: 
   html_document:
    keep_md: yes
---
## I Loading and preprocessing the data

### 1.1 Load the data

| Ensure your session is pointing to the directory with the csv file.
| Use read.csv command to load the data.
```{r}
A <- read.csv("activity.csv")
```

### 1.2 Preprocess the data
| Check structure of the loaded dataset with str. Review elements of the dataset using head.
| Convert the loaded data set into a data frame. Load package for coloring the graphs.

```{r}
str (A)
head(A)
A <- data.frame(A)
options(warn = -1)
require(RColorBrewer)
library(lattice)
```
## II What is the mean total number of steps taken per day?

### 2.1 Calculate the total number of steps taken per day 
| Use tapply to calculate the total number of steps per day. Omit missing values.
```{r}
AMOD <- tapply(A$steps, A$date, sum, na.rm = TRUE)
head (AMOD)
```
### 2.2 Make a histogram of the total number of steps taken each day 
| Call hist function to make a histogram of data from previous step
```{r histogram_2.3, fig.height = 6, fig.width = 6}

hist( AMOD, breaks = 10, lables = TRUE, ylim = c(0, 20), xlim = c(0, 25000), main="Histogram of # of steps per day", xlab="Number of Steps per day ", col = brewer.pal(5, "Set2"))
```
```{r phantom 1, echo = FALSE}

```

###2.3 Calculate and report the mean and median of the total number of steps taken per day 

```{r}
mean(AMOD)
median(AMOD)
```
| Mean of the total number of steps taken per day is `r round(mean(AMOD), digits = 2)`
| Median of the total number of steps taken per day is `r median(AMOD)`

## III What is the average daily activity pattern?
### 3.1 Make a time series plot of the 5-minute interval and the average number of steps taken

```{r}
BMOD <- tapply(A$steps, A$interval, mean, na.rm = TRUE)
plot(BMOD, type ="l", , main = "Avg steps across 5-minute intervals", xlab = "Interval Seq Number", ylab = "Avg. number of steps")
```
```{r phantom 2, echo = FALSE}

```
### 3.2 Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
| On average the maximum number of steps is `r round(max(BMOD))`. 
| The five minute interval number that contains the maximum number of steps is `r which(grepl(max(BMOD), BMOD))` th of the 288 total intervals.
| This corresponds to  `r (which(grepl(max(BMOD), BMOD))-1) * 5` minutes into the day. An inference is majority of the people going to work at 8:30 a.m. in the morning.


## IV Imputing missing values
### 4.1 Calculate and report the total number of missing values in the dataset 
| The total number of missing values is `r summary(A)[7,1]`.
```{r}
summary(A)
```
### 4.2 Devise a strategy for filling in all of the missing values in the dataset.
| Relplace missing value by mean of the number of steps in a given interval

```{r}
L <- BMOD[match(A$interval, names(BMOD))]
I.steps <- ifelse(is.na(A$steps), L, A$steps)
```
### 4.3 Create a new dataset that is equal to the original dataset but with the missing data filled in.
| Creating a new dataset by replacing the steps column by Imputed steps column in A
| Text 2

```{r}
NEWA <- A
NEWA$steps <- I.steps
names(NEWA)[1] <- "I.steps"
str(NEWA)
```
### 4.4 (a) Make a histogram of the total number of steps taken each day
| Use tapply to generate the column of data with total number of steps by day. Use hist to plot the histogram.

```{r}
IAMOD <- tapply(NEWA$I.steps, NEWA$date, sum)
hist( IAMOD, breaks = 10, labels = TRUE, ylim = c(0, 20), xlim = c(0, 25000), main="Histogram of # of steps per day", xlab="Number of Steps per day ", col = brewer.pal(5, "Set2"))
```
```{r phantom 3, echo = FALSE}

```
### 4.4 (b) Calculate and report the mean and median of total number of steps taken per day
```{r}
mean(IAMOD)
median(IAMOD)
```
| Mean of the total number of imputed steps taken per day is `r round(mean(IAMOD), digits = 2)`
| Median of the total number of imputed steps taken per day is `r median(IAMOD)`
|
| The difference between the means (total steps, total imputed steps) is `r round(mean(AMOD), digits = 2) - round(mean(IAMOD), digits = 2)`
| The difference between the medians (total steps, total imputed steps) is `r median(AMOD) - median(IAMOD)`

## V Weekdays and weekends activity patterns
### 5.1 Create a new factor variable for weekdays and weekends
| Use ifelse function to compute a vector of weekday/weekend. Column bind this vector to NEWA dataset.
```{r}
day <- weekdays(as.Date(NEWA$date))
type <- ifelse(day %in% c("Saturday", "Sunday"), "weekend", "weekday")
M1 <- cbind(NEWA, type)
str(M1)
head(M1)
```
### 5.2 Make a panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 
| Subset data in M1 by weekend and weekday.
```{r}
M1WDAY <- subset(M1, type == "weekday")
head(M1WDAY)
M1WEND <- subset(M1, type == "weekend")
head(M1WEND)
```
| Use tapply to create sums by interval for each of the above subsets,
| Use base plot function with type = l to create time series plots.
```{r fig.height = 10}
par(mfrow = c(2,1))
WDAYMOD <- tapply(M1WDAY$I.steps, M1WDAY$interval, mean)
plot(WDAYMOD, type ="l", main = "Weekday - Avg steps across 5-minute intervals", xlab = "Interval Seq Number", ylab = "Avg. number of steps")
WENDMOD <- tapply(M1WEND$I.steps, M1WEND$interval, mean)
plot(WENDMOD, type ="l", main = "Weekend - Avg steps across 5-minute intervals", xlab = "Interval Seq Number", ylab = "Avg. number of steps")
```
```{r phantom 4, echo = FALSE}

```
| OBSERVATION:
| WEEKDAYS:
| The five minute interval number that contains the maximum number of steps is `r which(grepl(max(WDAYMOD), WDAYMOD))` th of the 288 total intervals. This corresponds to  `r (which(grepl(max(WDAYMOD), WDAYMOD))-1) * 5` minutes past the hour. An inference from this data is that most people reach work place by 8:30 a.m.
|
| WEEKENDS:
| On weekends we see several maximums all scross the day. This shows folks are active all day during the weekend.
|
| NOTES:
| Please note - I attempted this with lattice panel. The plot with lattice has noise.
```{r fig.height = 6}
#M2 <- transform(M1, type = factor(type))
xyplot(I.steps ~ interval | type, data = M1, type = "l", layout= c(1,2))
```
```{r phantom 5, echo = FALSE}

```

